Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-29 paper-1/manuscript.tex
SSA disambiguation

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in our model. \textbf{a} Sample network target response intensity by trial type. Each horizontal line of the plot summarizes the network response in one target trial in terms of spikes per integration time step. Target trials are taken from the std (400), msc (100), and dev (100) sequences, and sorted by sequence time (earliest trials are at the bottom of each section). \textbf{b} Response summary, showing average network-wide spike counts in target trials in each sequence. The underlying data are the same as in panel a. Note how the dev response is very large, followed by an intermediate msc response and a low std response. \textbf{c} Mismatch, adaptation, and deviance detection indices (see Equation \ref{eqn-ddi}) across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points

We define deviance detection as a network's ability to pick out unexpected inputs from an otherwise homogeneous or otherwise unsurprising stream of inputs. Following related research \citep{Kubota2021-dx,Harms2014-ah,Jacobsen2001-sc}, we were careful to exclude effects due to stimulus identity (i.e., avoiding direct comparisons between different stimuli) or due to adaptation to frequent stimulus presentation (i.e., avoiding direct comparisons of A in the dev sequence to A in the std sequence). In other words, we used the msc sequence as a neutral baseline where the target stimulus was presented no more often than in the dev sequence, but no expectation of a competing stimulus B could be formed. To quantify deviance detection, we compared the mean number of spikes $R^A$ fired across the network in response to target stimulus A in the dev and msc sequences, defining a deviance detection index (DDI) as
\begin{equation}
    DDI = \frac{R^A_{dev} - R^A_{msc}}{R^A_{dev} + R^A_{msc}} \label{eqn-ddi}
\end{equation}
We define stimulus-specific adaptation (SSA, $R^A_{msc} - R^A_{std}$) and mismatch ($R^A_{dev} - R^A_{std}$) indices analogously, capturing respectively the effect of adaptation, and of both adaptation and deviance detection together.

@@

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the DDI and related indices mentioned above for each network, and show them in a summary plot in Figure \ref{fig2}c. We found that our model reliably produced positive mismatch indices (T = 6.67e+03, p = 9.26e-16, median 0.253; one-sided Wilcoxon signed-rank test, n = 120) while both SSA index (T = 6.19e+03, p = 9.76e-12, median 0.135) and DDI (T = 5.97e+03, p = 4.75e-10, median 0.114) were also positive, despite some outliers. This shows that our model was well behaved and suitable for an investigation of deviance detection.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA). \textbf{b} Adaptation indices $<$msc - std$>$ across all networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics. \textbf{c} Deviance detection indices $<$dev - msc$>$ across all networks and stimuli

Having confirmed the presence of deviance detection in our model, we then turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. In the remainder of the paper, we will proceed to detail the processes leading to deviance detection on a sample network, then confirm the highlighted trends statistically with reference to the full set of 120 networks and stimuli. The sample network was chosen to be roughly representative of the population as follows: We assessed each network's average response to stimulation with A and B in the msc sequence and calculated z-scores with respect to the population. We then focused on networks whose absolute z-score was less than 1 for both A and B responses, and hand-picked a sample that exhibited both an increased DDI in STD+TA over TA only, and positive SSA indices in both STD+TA and TA only conditions.

@@

The SSA index (Figure \ref{fig3}b) was exactly 0 as expected in the ablation with no plasticity, and positive with any plasticity applied (STD only: T = 4.97e+03, p = 0.000222, median 0.0039; TA only: T = 5.99e+03, p = 3.3e-10, median 0.097; STD+TA: T = 6.19e+03, p = 9.76e-12, median 0.135; one-sided Wilcoxon signed-rank tests, n = 120). Since the SSA index measures the degree of adaptation to frequent stimulus presentation, this indicates that both plasticity mechanisms were capable of imbuing the networks with some level of history dependence.

@@

We had good reasons for this assumption: Firstly, referring back to Figure \ref{fig3}, notice that the ablated model with STD only produced only a positive SSA index (i.e., the target response in msc was greater than in std), but not a significant DDI (i.e., no difference between target dev and msc responses). Secondly, we note that updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B could lead to a build-up of depression strong enough to suppress its own response.
To show that the network does not carry information beyond the most recent trial in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in our model. \textbf{a} Sample network target response intensity by trial type. Each horizontal line of the plot summarizes the network response in one target trial in terms of spikes per integration time step. Target trials are taken from the std (400), msc (100), and dev (100) sequences, and sorted by sequence time (earliest trials are at the bottom of each section). \textbf{b} Response summary, showing average network-wide spike counts in target trials in each sequence. The underlying data are the same as in panel a. Note how the dev response is very large, followed by an intermediate msc response and a low std response. \textbf{c} Mismatch, repetition suppression, and deviance detection indices (see Equation \ref{eqn-ddi}) across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points

We define deviance detection as a network's ability to pick out unexpected inputs from an otherwise homogeneous or otherwise unsurprising stream of inputs. Following related research \citep{Kubota2021-dx,Harms2014-ah,Jacobsen2001-sc}, we were careful to exclude effects due to stimulus identity (i.e., avoiding direct comparisons between different stimuli) or due to adaptation to frequent stimulus presentation (i.e., avoiding direct comparisons of A in the dev sequence to A in the std sequence). In other words, we used the msc sequence as a neutral baseline where the target stimulus was presented no more often than in the dev sequence, but no expectation of a competing stimulus B could be formed. To quantify deviance detection, we compared the mean number of spikes $R^A$ fired across the network in response to target stimulus A in the dev and msc sequences, defining a deviance detection index (DDI) as
\begin{equation}
    DDI = \frac{R^A_{dev} - R^A_{msc}}{R^A_{dev} + R^A_{msc}} \label{eqn-ddi}
\end{equation}
We define related indices, contrasting the responses in msc against std, and in dev against std, analogously. Since contrasts of both kinds have been called "stimulus-specific adaptation" or SSA in the literature \citep{Kubota2021-dx,Taaseh2011-gg}, we chose to avoid such ambiguity, using instead the term "repetition suppression" (RSI, $R^A_{msc} - R^A_{std}$) to refer to the reduction in response induced mainly by a difference in presentation frequency while controlling for context, and the term "mismatch" (MMI, $R^A_{dev} - R^A_{std}$) to refer to the uncontrolled contrast, which captures a combination of repetition suppression and deviance detection.

@@

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the DDI and related indices mentioned above for each network, and show them in a summary plot in Figure \ref{fig2}c. We found that our model reliably produced positive mismatch indices (T = 6.67e+03, p = 9.26e-16, median 0.253; one-sided Wilcoxon signed-rank test, n = 120) while both RSI (T = 6.19e+03, p = 9.76e-12, median 0.135) and DDI (T = 5.97e+03, p = 4.75e-10, median 0.114) were also positive, despite some outliers. This shows that our model was well behaved and suitable for an investigation of deviance detection.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA). \textbf{b} Repetition suppression indices $<$msc - std$>$ across all networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics. \textbf{c} Deviance detection indices $<$dev - msc$>$ across all networks and stimuli

Having confirmed the presence of deviance detection in our model, we then turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. In the remainder of the paper, we will proceed to detail the processes leading to deviance detection on a sample network, then confirm the highlighted trends statistically with reference to the full set of 120 networks and stimuli. The sample network was chosen to be roughly representative of the population as follows: We assessed each network's average response to stimulation with A and B in the msc sequence and calculated z-scores with respect to the population. We then focused on networks whose absolute z-score was less than 1 for both A and B responses, and hand-picked a sample that exhibited both an increased DDI in STD+TA over TA only, and positive RSI in both STD+TA and TA only conditions.

@@

The RSI (Figure \ref{fig3}b) was exactly 0 as expected in the ablation with no plasticity, and positive with any plasticity applied (STD only: T = 4.97e+03, p = 0.000222, median 0.0039; TA only: T = 5.99e+03, p = 3.3e-10, median 0.097; STD+TA: T = 6.19e+03, p = 9.76e-12, median 0.135; one-sided Wilcoxon signed-rank tests, n = 120). Since the RSI measures the degree of adaptation to frequent stimulus presentation, this indicates that both plasticity mechanisms were capable of imbuing the networks with some level of history dependence.

@@

We had good reasons for this assumption: Firstly, referring back to Figure \ref{fig3}, notice that the ablated model with STD only produced only a positive RSI (i.e., the target response in msc was greater than in std), but not a significant DDI (i.e., no difference between target dev and msc responses). Secondly, we note that updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B could lead to a build-up of depression strong enough to suppress its own response.
To show that the network does not carry information beyond the most recent trial in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial.
